---
title: 'outlier detection'
---


```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```
```{r}
setwd('C:/Users/Elena/Desktop/Elena/Polimi/MAGISTRALE/Nonparametric statistics/Progetto/github repository/ALZHEIMER_prognonpa/Elena')
dataset_xsectional <- read.csv("oasis_cross-sectional.csv", header = T)
dataset_longitudinal <- read.csv("oasis_longitudinal.csv", header = T)
```
```{r}
library(DepthProc)
library(aplpack)
```



To visualize the outliers in this context we retort to a bagplot matrix:

```{r}
bagplot_matrix <- aplpack::bagplot.pairs(dataset_longitudinal[,c(8,9,11,13,14,15)], main = 'Pairs of bagplots')  # bagplot dimension by dimension
```

Using this bagplot we want to try to understan which comparisons are reasonable to find outliers using the depth measures.


Indeed the outlier detection becomes more difficult when the dimension increases, as

-   some outliers may be wrongly flagged as genuine points
-   some good points may be wrongly flagged as outliers

These phenomena are respectively denoted as **masking** and **swamping**, we will cover it in details during the robust statistics analysis




```{r}
depthContour(
  data.frame(dataset_longitudinal$Age, dataset_longitudinal$MMSE),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)
```


```{r}
bags1 <- bagplot(data.frame(dataset_longitudinal$Age, dataset_longitudinal$MMSE), xlab = "Age", ylab = "MMSE")

```
```{r}
outlying_obs1 <- bags1$pxy.outlier

outlying_obs1

which(dataset_longitudinal$Age==outlying_obs1[,1] & dataset_longitudinal$MMSE==outlying_obs1[,2])
```





```{r}
depthContour(
  data.frame(dataset_longitudinal$EDUC, dataset_longitudinal$MMSE),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags2 <- bagplot(  data.frame(dataset_longitudinal$EDUC, dataset_longitudinal$MMSE), xlab = "EDUC", ylab = "MMSE")
```

```{r}
outlying_obs2 <- bags2$pxy.outlier

outlying_obs2

which(dataset_longitudinal$EDUC==outlying_obs2[,1] & dataset_longitudinal$MMSE==outlying_obs2[,2])
```


```{r}
depthContour(
  data.frame(dataset_longitudinal$eTIV, dataset_longitudinal$MMSE),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags3 <- bagplot(data.frame(dataset_longitudinal$eTIV, dataset_longitudinal$MMSE), xlab = "eTIV", ylab = "MMSE")
```

```{r}
outlying_obs3 <- bags3$pxy.outlier

outlying_obs3

which(dataset_longitudinal$eTIV==outlying_obs3[,1] & dataset_longitudinal$MMSE==outlying_obs3[,2])
```

```{r}
depthContour(
  data.frame(dataset_longitudinal$nWBV, dataset_longitudinal$MMSE),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags4 <- bagplot(data.frame(dataset_longitudinal$nWBV, dataset_longitudinal$MMSE), xlab = "nWBV", ylab = "MMSE")
```

```{r}
outlying_obs4 <- bags4$pxy.outlier

outlying_obs4

which(dataset_longitudinal$nWBV==outlying_obs4[,1] & dataset_longitudinal$MMSE==outlying_obs4[,2])
```


```{r}
depthContour(
  data.frame(dataset_longitudinal$ASF, dataset_longitudinal$MMSE),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags5 <- bagplot(data.frame(dataset_longitudinal$ASF, dataset_longitudinal$MMSE), xlab = "ASF", ylab = "MMSE")
```

```{r}
outlying_obs5 <- bags5$pxy.outlier

outlying_obs5


which(dataset_longitudinal$ASF==outlying_obs5[,1] & dataset_longitudinal$MMSE==outlying_obs5[,2])
```


```{r}
depthContour(
  data.frame(dataset_longitudinal$Age,dataset_longitudinal$EDUC ),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags6 <- bagplot(data.frame(dataset_longitudinal$Age,dataset_longitudinal$EDUC ), xlab = "Age", ylab = "EDUC")
```
```{r}
outlying_obs6 <- bags6$pxy.outlier

outlying_obs6

which(dataset_longitudinal$Age==outlying_obs6[,1] & dataset_longitudinal$EDUC==outlying_obs6[,2])
```


We can notice that these three persons have an eduction really lower wth respect to the others.




```{r}
depthContour(
  data.frame(dataset_longitudinal$eTIV, dataset_longitudinal$EDUC),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags7 <- bagplot(data.frame(dataset_longitudinal$eTIV, dataset_longitudinal$EDUC), xlab = "eTIV", ylab = "EDUC")
```
```{r}
outlying_obs7 <- bags7$pxy.outlier

outlying_obs7

which(dataset_longitudinal$eTIV==outlying_obs7[,1] & dataset_longitudinal$EDUC==outlying_obs7[,2])
```


```{r}
depthContour(
  data.frame(dataset_longitudinal$nWBV, dataset_longitudinal$EDUC),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags8 <- bagplot(data.frame(dataset_longitudinal$nWBV, dataset_longitudinal$EDUC), xlab = "nWBV", ylab = "EDUC")
```
```{r}
outlying_obs8 <- bags8$pxy.outlier

outlying_obs8

which(dataset_longitudinal$nWBV==outlying_obs8[,1] & dataset_longitudinal$EDUC==outlying_obs8[,2])
```


```{r}
depthContour(
  data.frame(dataset_longitudinal$ASF, dataset_longitudinal$EDUC),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags9 <- bagplot(data.frame(dataset_longitudinal$ASF, dataset_longitudinal$EDUC), xlab = "ASF", ylab = "EDUC")
```


```{r}
outlying_obs9 <- bags9$pxy.outlier

outlying_obs9

which(dataset_longitudinal$ASF==outlying_obs9[,1] & dataset_longitudinal$EDUC==outlying_obs9[,2])
```

```{r}
depthContour(
  data.frame(dataset_longitudinal$Age, dataset_longitudinal$eTIV),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags10 <- bagplot(data.frame(dataset_longitudinal$Age, dataset_longitudinal$eTIV), xlab = "Age", ylab = "eTIV")
```


```{r}
outlying_obs10 <- bags10$pxy.outlier

outlying_obs10

which(dataset_longitudinal$Age==outlying_obs10[,1] & dataset_longitudinal$eTIV==outlying_obs10[,2])
```



The patient OAS2_0048 at line 101 results as anomalous in the fourth visit for MMSE vs Age.

The patient OAS2_0048 at line 102 results as anomalous in the fifth visit for both MMSE vs Age, and MMSE vs the last 3 columns (that are dependents).

The patient	OAS2_0040 at lines 79, 80, 81 anomalous in all the 3 visits for EDUC vs Age and vs the last 3 columns.

The patient OAS2_0066 at line 140 results as anomalous in the first visit only for Age vs eTIV (!attention: non per le altre due colonne con cui etiv è dipendente! -> può essere il valore di etiv sbagliato nella riga 140? e soprattutto non LL visita successiva)




Now we'll do the same with the complete training dataset considering only the first visit:

```{r}
dataset_train <- read.csv("train.csv", header = T)
dataset_train <- dataset_train[,-1]
dataset_train$L <- ifelse(dataset_train$label == 'Dem', 1, 0)  # dummy variable for demented/nondemented
head(dataset_train)
```





To visualize the outliers in this context we retort to a bagplot matrix:

```{r}
bagplot_matrix_train <- aplpack::bagplot.pairs(dataset_train[,c(-1,-7,-8, -9)])
```

it seems to have problems only Age with eTIV and nWBV


Age vs eTIV:

```{r}
depthContour(
  data.frame(dataset_train$Age, dataset_train$eTIV),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags1_train <- bagplot(data.frame(dataset_train$Age, dataset_train$eTIV), xlab = "Age", ylab = "eTIV")

```
```{r}
outlying_obs1_train <- bags1_train$pxy.outlier

outlying_obs1_train

which(dataset_train$Age==outlying_obs1_train[1,1] & dataset_train$eTIV==outlying_obs1_train[1,2])
which(dataset_train$Age==outlying_obs1_train[2,1] & dataset_train$eTIV==outlying_obs1_train[2,2])
```


Age vs nWBV:

```{r}
depthContour(
  data.frame(dataset_train$Age, dataset_train$nWBV),
  depth_params = list(method = 'Tukey'),
  points = TRUE,
  colors = colorRampPalette(c('white', 'navy')),
  levels = 10,
  pdmedian = F,
  graph_params = list(cex=.01, pch=1),
  pmean = F
)

bags2_train <- bagplot(data.frame(dataset_train$Age, dataset_train$nWBV), xlab = "Age", ylab = "nWBV")

```
```{r}
outlying_obs2_train <- bags2_train$pxy.outlier

outlying_obs2_train

which(dataset_train$Age==outlying_obs2_train[,1] & dataset_train$nWBV==outlying_obs2_train[,2])
```


In the dataset 'train.csv':
-at line 175 there is a patient anomalous both for Age vs eTIV and Age vs nWBV: he is 33 y.o.
-at line 145 there is a patient anomalous only for Age vs eTIV: he is 39 y.o.



