---
title: "SURVIVAL CORRETTO"
output: html_document
date: '2022-06-28'
---
---
title: "Lab 13 - Survival analysis"
date: 2021/11/19
author: "Nonparametric statistics ay 2021/2022"
output:
  
  html_document: 
    df_print: paged
  pdf_document: default
  html_notebook: 
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
options(rgl.debug= T)
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
ggplot2::theme_set(ggplot2::theme_bw())
```


## Loading necessary libraries

```{r message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(dplyr) 
library(ggplot2)
library(knitr)
library(broom)
```
```{r}
ind <- dataset[dataset$Group=='Converted'& dataset$Visit==1,]
```

PREPARAZIONE DEL DATASET

```{r}
setwd("C:/Users/franc/Desktop/NONPA/PROGETTO")

library(stringr)
dataset_old <- read.csv("oasis_longitudinal.csv")
head(dataset_old)
colnames(dataset_old)

## DEMENTI  
dataset_old$ID = str_sub(dataset_old$Subject.ID,-3) 
dementi_prima <- dataset_old[dataset_old$Group=='Demented'& dataset_old$Visit==1,]
dementi_prima$status <- 2
dementi_prima$group_new <- 'Demented'
##NON DEMENTI
nondementi_last <- dataset_old[dataset_old$Group=='Nondemented',]
nondementi_last <-nondementi_last%>%group_by(ID)%>% mutate(max_visit=max(Visit))%>%ungroup()%>% filter(Visit==max_visit)
length(unique(nondementi_last$ID))
nondementi_last$status<-1
nondementi_last$group_new <- 'Nondemented'
## CONVERTED

converted_first <- dataset_old[dataset_old$Group=='Converted'&dataset_old$CDR >0,]
converted_first <-converted_first%>%group_by(ID)%>% mutate(min_visit=min(Visit))%>%ungroup()%>% filter(Visit==min_visit)
converted_first$group_new <- 'Demented'
converted_first$status <- 2
#ALL TOGETHER 
dataset <- rbind(nondementi_last[,-17],dementi_prima,converted_first[,-17])

dataset$time_y <- dataset$MR.Delay / 365
dataset$status_fact <- factor(dataset$Group)
colnames(dataset) <- c( "Subject.ID",  "MRI.ID"   ,   "Group"   ,    "Visit"   ,    "time" , "M.F"       ,  "Hand" , "age"     ,    "EDUC"   ,     "SES"  ,       "MMSE"     ,   "CDR"     ,    "eTIV"    ,    "nWBV"   ,   "ASF"  ,       "ID"     ,     "status"   ,   "group_new"  , "time_y"    ,  "status_fact")

```


```{r}
ind_ <- which(dataset$ID=='131')
x11()
ggplot(data=dataset[-146,],aes(x=ID,y=time_y)) + 
  geom_bar(stat='identity',width=0.2) +
  geom_point(aes(color=status_fact,shape=status_fact),size=6) +
  coord_flip()


```


## Kaplan-Meier estimator for survival curve
 $$ S(t) $$ 
```{r}
fit <- survfit(Surv(dataset$time, status==2) ~ 1, data = dataset)
# status==1 -> when the event occurred

summary(fit)

```


```{r}
kable(head(tidy(fit),20))
```

The median survival times represents the time at which the survival
probability, S(t), is 0.5.

```{r,warning=FALSE}
surv_median(fit)
# median = 1331 -> median time of event
```


### Kaplan-Meier plots

```{r}
ggsurvplot(fit,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), # Change ggplot2 theme
           break.time.by=360,
           title="Kaplan-Meier Curve for Alzheimer Survival")
```



```{r}
cumulative_incidence <- 1 - fit$surv
```


```{r}
ggsurvplot(fit,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), # Change ggplot2 theme
           break.time.by=90,
           fun='event',
           title="Cumulative Incidence Curve for Alzheimer Survival")
```


Nelson-Aalen cumulative hazard rate estimator 

```{r}
H <- fit$cumhaz
```


```{r}
ggsurvplot(fit,
           risk.table = TRUE, # Add risk table
           ggtheme = theme_bw(), # Change ggplot2 theme
           break.time.by=360,
           fun='cumhaz',
           title="Cumulative Hazard Curve for Alzheimer Survival")
```

### Kaplan-Meier Curves between groups: GENDER

We want to consider now the gender groups and investigate if there is a
difference in terms of survival among the two groups.
$$S(t) \sim Sex $$



```{r}
dataset$sex <- 0
dataset$sex[which(dataset$M.F=='F')] <-1
fit.sex <- survfit(Surv(time, status) ~ sex, data=dataset)
surv_median(fit.sex)
ggsurvplot(fit.sex, conf.int = T,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), # Change ggplot2 theme
           break.time.by=360,
           legend.title="SEX",  legend.labs=c('M','F'), 
           palette=c("darkblue","red"), 
           title="Kaplan-Meier Curves by gender class for Alzheimer Survival")
```



It looks like there's some differences in the curves between gender class patients. Is there statistical evidence for that difference?

```{r}

log_rank_test <- survdiff(Surv(time, status) ~ sex,data=dataset)
log_rank_test  
```

$p=0.01$, the difference in survival between the gender class is significant


## Hazard Ratio

```{r}
hazard_ratio <- (log_rank_test$obs[1]/log_rank_test$exp[1])/(log_rank_test$obs[2]/log_rank_test$exp[2])
hazard_ratio #

```
HR< 1 -> group 1 protective factor ( less risk)
HR= 1 -> similar risk
HR> 1 -> group 1 risk factor (more risk)

$HR_{M,F} = 1.570315> 1$ 

being a male is a risk factor risk



## Cox model

 Let us start by
plotting a histogram showing the distribution of age.

```{r}
attach(dataset)
hist(dataset[sex=='0'& Group=='Converted',]$age, xlab='Age [years]', main='Histogram of age of Converted Male in Alzheimer Data', col="darkblue",bin=10)
hist(dataset[sex=='1'& Group=='Converted',]$age, xlab='Age [years]', main='Histogram of age of Converted Female in Alzheimer Data', col="red")
hist(dataset$age, xlab='Age [years]', main='Histogram of  age in Alzheimer Data')
summary(dataset[sex=='1',]$age)
hist(dataset[dataset$Group=='Converted',]$age, xlab='Age [years]', main='Histogram of age of Converted patient in Alzheimer Data', col=cols[2])
```


Let us consider the continuous variable age and fit a univariate Cox
regression model

```{r}
cox.age <- coxph(Surv(time_y, status) ~ age, data = dataset)
cox.age
```

```{r}
summary(cox.age)
```
```{r}
library(RColorBrewer)
cols <- brewer.pal(3, "RdPu")
plot(survfit(cox.age, data=dataset), 
     col=cols[3], lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability')
grid()
abline(h=0.5, l='-')
```

# age and sex

```{r}
cox.age.sex <- coxph(Surv(time_y, status) ~ age + sex, data = dataset)
cox.age.sex
```
```{r}
summary(cox.age.sex)
```


```{r}
cols=  brewer.pal(3, "RdPu")
library(RColorBrewer)
plot(survfit(cox.age.sex, data=dataset), 
     col=cols[3], lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability wrt Age and Gender')
grid()
```

## Multivariate Cox regression

We want now to describe how different factors jointly impact on
survival. To answer to this question, we will perform a multivariate Cox
regression analysis with covariates age, sex, Karnofsky performance
score rated by physician and weight loss. Check if you categorical
covariates are considered factors:

```{r}
glimpse(dataset)

dataset$sex <- as.factor(dataset$M.F)

```

Fit the Cox's regression model:

```{r}
mod.cox <- coxph(Surv(time, status) ~ age + sex + EDUC + MMSE, data =  dataset)
summary(mod.cox)
```


### Visualizing Hazard ratios

You can visualize Hr and its CIs using the `ggforest()` function of
package survminer:

```{r}
ggforest(mod.cox, data=dataset)
```
Because the confidence interval for HR includes 1, these
results indicate that age makes a smaller contribution to the difference
in the HR after adjusting for the other covariates.

Plot the baseline survival function $S_0(t)$

```{r}
plot(survfit(mod.cox, data=dataset), 
     col="darkorange2", lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability')
grid()
```
```{r}
mod.cox_age_sex <- coxph(Surv(time, status) ~ age + sex , data =  dataset)
summary(mod.cox_age_sex)
```
```{r}
plot(survfit(mod.cox_age_sex, data=dataset), 
     col=cols[3], lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability')
grid()
```

```{r}
mod.cox_MMSE <- coxph(Surv(time, status) ~ MMSE , data =  dataset)
summary(mod.cox_MMSE)
```
```{r}
plot(survfit(mod.cox_MMSE, data=dataset), 
     col=cols[3], lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability')
grid()
```

```{r}
mod.cox_ALL <- coxph(Surv(time, status) ~ age + sex + EDUC + MMSE+ SES + eTIV, data =  dataset)
summary(mod.cox_ALL)
```

```{r}
plot(survfit(mod.cox_ALL, data=dataset), 
     col=cols[3], lwd=2, lty=1,
     xlab='Time ', ylab='Survival Probability',
     main='Baseline estimated survival probability')
grid()
```

alla fine il più significativo è sempre MMSE
### Cox Model Assumptions and Goodness of fit


When used inappropriately, statistical models may give rise to
misleading conclusions. Therefore, it is important to check that a given
model is an appropriate representation of the data.

A first graphical option to check the goodness of fit is to check if the
Martingale Residuals

$$
M_i=\delta_i-H(t_i, \mathbf{X}_i, \boldsymbol{\beta}), \quad n=1,\ldots,N
$$

with $\delta_i$ 0-1 function indicating whether the $i$-th unit has
experienced the event (1 if present, 0 otherwise) and

$$
H(t_i, \mathbf{X}_i, \boldsymbol{\beta})=-\log \left[\hat{S}\left(t_i,\mathbf{X}_i, \boldsymbol{\beta} \right)\right]
$$

have $0$ mean along time. Recall that in a Cox model for each patient
$i$ the corresponding survival function is estimated with $$
\hat{S}\left(t ,\mathbf{X}_i, \boldsymbol{\beta}\right)=\left[\hat{S}_{0}(t)\right]^{\exp \left(\boldsymbol{X}_{i}^{T} \boldsymbol{\beta}\right)}$$

where

$$\hat{S_{0}}(t)=\prod_{j: t_{j}^{*}<t}\left(1-\frac{1}{\sum_{k \in R\left(t_{j}^{*}\right)} \exp \left(\boldsymbol{X}_{k}^{T} \hat{\boldsymbol{\beta}}\right)}\right)$$



```{r}
ggcoxdiagnostics(mod.cox, type = "martingale", labels= 'age + sex + EDUC + MMSE')
```
```{r}
ggcoxdiagnostics(mod.cox_MMSE, type = "martingale" ,labels='MMSE')
```
```{r}
ggcoxdiagnostics(mod.cox_ALL, type = "martingale" ,labels='ALL')
```

```{r}
ggcoxdiagnostics(mod.cox_age_sex, type = "martingale",labels= 'age sex')
```


Sometimes, martingale residuals are difficult to be interpreted. The
deviance residual is a normalized transform of the martingale residual:

$$ \hat{D}_{i}=\operatorname{sign}\left(M_{i}\right) \sqrt{-2\left[M_{i}+\delta_{i} \log \left(\delta_{i}-M_{i}\right)\right]} \quad=1,\ldots,N$$

These residuals should be roughly symmetrically distributed about zero
with a standard deviation of 1.

-   Positive values correspond to individuals that "died too soon"
    compared to expected survival times.
-   Negative values correspond to individual that "lived too long".
-   Very large or small values are outliers, which are poorly predicted
    by the model.

It is also possible to check outliers by visualizing the deviance
residuals. Example of deviance residuals:

```{r}
ggcoxdiagnostics(mod.cox, type = "deviance")
```
```{r}
ggcoxdiagnostics(mod.cox_MMSE, type = "deviance" ,labels='MMSE')
```
The pattern doesn't look fairly symmetric around $0$, not a good fit
```{r}
ggcoxdiagnostics(mod.cox_ALL, type = "deviance" ,labels='ALL')
```

```{r}
ggcoxdiagnostics(mod.cox_age_sex, type = "deviance",labels= 'age sex')
```

A second graphical option could be to use the Schoenfeld residuals to
examine model fit and detect outlying covariate values. Shoenfeld
residuals represent the difference between the observed covariate and
the expected given the risk set at that time. They should be flat,
centered about zero. In principle, the Schoenfeld residuals are
independent of time. A plot that shows a non-random pattern against time
is evidence of violation of the PH assumption.

```{r}
ggcoxdiagnostics(mod.cox, type = "schoenfeld")
```
```{r}
ggcoxdiagnostics(mod.cox_MMSE, type = "schoenfeld")
```
```{r}
ggcoxdiagnostics(mod.cox_ALL, type = "schoenfeld" ,labels='ALL')
```

```{r}
ggcoxdiagnostics(mod.cox_age_sex, type = "schoenfeld",labels= 'age sex')
```


Another graphical method for checking proportional hazards is to plot
$log(-log(KM(t)))$ vs. $t$ or $log(t)$ and look for parallelism. This
can be done only for categorical covariates.

We consider the KM estimators for sex variable:

```{r}
sex.km <- survfit(Surv(time, status) ~ sex, data = dataset)
```

We plot $log(-log(KM(t)))$ using option `fun='cloglog'` in
`plot.survfit()`

```{r}
plot(sex.km, fun='cloglog', 
     col=c("deeppink2","dodgerblue2"), lwd=2, lty=1,
     ylab="log(-log(Survival Probability))")
grid()
legend('topleft', c("Female", "Male"),
       lty=c(1,1), lwd=c(2,2), col=c("deeppink2","dodgerblue2"))
```

Curves seem to be NOT parallel -\> PH assumption seems NOT satisfied for
gender.

The function `cox.zph()` in the survival package provides a convenient
solution to test the proportional hazards assumption for each covariate
included in a Cox regression model fit.

For each covariate, the function cox.zph() correlates the corresponding
set of scaled Schoenfeld residuals with time, to test for independence
between residuals and time. Additionally, it performs a global test for
the model as a whole.

The proportional hazard assumption is supported by a non-significant
relationship between residuals and time, and refused by a significant
relationship.

Test for PH using scaled Schoenfeld test for PH

-   H0: Hazards are proportional
-   H1: Hazards are NOT proportional

cox.zph() return tests for each X and for the global model

```{r}
test.ph <- cox.zph(mod.cox)
test.ph
```
From the output above, the global test is statistically significant (small pvalue, reject H0).
Therefore, we can not assume the proportional hazards. In particular,
the test for ph.MMSE is highly significant.

```{r}
test.ph_ALL <- cox.zph(mod.cox_ALL)
test.ph_ALL
```
same as before 
```{r}
test.ph_MMSE <- cox.zph(mod.cox_MMSE)
test.ph_MMSE
```

```{r}
test.ph_age_sex <- cox.zph(mod.cox_age_sex)
test.ph_age_sex
```
age is higly significant

Plot the scaled schoenfeld residuals:


```{r}
ggcoxdiagnostics(mod.cox, type = "scaledsch")
```

```{r}
ggcoxdiagnostics(mod.cox_MMSE, type = "scaledsch")
```

```{r}
ggcoxdiagnostics(mod.cox_ALL, type = "scaledsch")
```

```{r}
ggcoxdiagnostics(mod.cox_age_sex, type = "scaledsch")
```

So... What do we do? As a very basic primer on Survival Analysis we will
not have time to thoroughly cover all possible solutions, we will
(briefly) focus on stratification.

## Stratified Cox Model

Sometimes the proportional hazard assumption is violated for some
covariate. In such cases, it is possible to stratify taking this
variable into account and use the proportional hazards model in each
stratum for the other covariates. We include in the model predictors
that satisfy the proportional hazard assumption and remove from it the
predictor that is stratified.

Now, the subjects in the $k$-th stratum have an arbitrary baseline
hazard function $h_{0k}(t)$ and the effect of other explanatory
variables on the hazard function can be represented by a proportional
hazards model in that stratum: $$h_{k}(t|X) = h_{0k}(t) \exp(\beta^TX)$$
with $k=1,\ldots,K$ levels of the variable that is stratified.

In the Stratified Proportional Hazards Model the regression coefficients
are assumed to be the same for each stratum although the baseline hazard
functions may be different and completely unrelated. The model may seem
complex, but it is entirely straightforward in the likelihood framework,
as we can simply combine likelihoods across strata (i.e., we multiply
each strata-wise contribution). This is easily accomplished in R by
using the `strata()` argument:

```{r}
mod.cox.strata <- coxph(Surv(time, status) ~ age + sex + strata(MMSE) , data =  dataset)
summary(mod.cox.strata)
```

Test for PH assumption

```{r}
test.ph.strata <- cox.zph(mod.cox.strata)
test.ph.strata
```
 
PH assumptions are satisfied for all variables and for the global model.

```{r}
mod.cox.strata_ALL <- coxph(Surv(time, status) ~  age + sex + EDUC + strata(MMSE)+ SES + eTIV, data =  dataset)
summary(mod.cox.strata_ALL)
```

Test for PH assumption

```{r}
test.ph.strata_ALL <- cox.zph(mod.cox.strata_ALL)
test.ph.strata_ALL
```



###PREDICTION


```{r}
attach(train)
train$nWBV<- as.numeric(train$nWBV)
fem_pt <- DepthProc::depthMedian(as.matrix(train[M.F=='F'&label=='Nondem',3:7]),list(method='Tukey'))
unique(label)
train$label <-  as.factor(train$label, levels=c('Nondem','Dem'))
mod.cox_ALL <- coxph(Surv(time, status) ~ age + sex + EDUC + MMSE+ SES + eTIV, data =  dataset)
summary(mod.cox_ALL)
new_df <- with(dataset,
               data.frame(age = c(50,65,80) )#consider ages equal to 50, 65 and 80
)

#estimate survival:
fit.new <- survfit(cox, newdata = new_df)
fit.new
```

